Temporal ordering and registration of imaging data in cross--sectional studies of development

### 2014-03-24 image_analysis_paper/pnas_draft.tex
draft of pnas paper

\documentclass[11pt]{article}

\usepackage{amsmath}
\usepackage{amssymb}
\DeclareMathOperator*{\argmin}{arg\,min}

\usepackage{graphicx}
\usepackage{subcaption}
\graphicspath{ {paper_figures/} }

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{fig2a}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{fig2b}
\caption{}
\end{subfigure}
\caption{} 
\label{fig:fluorescent_images}
\end{figure}

\section{Techniques}

\subsection{Registration Using Angular Synchronization} \label{subsec:angsynch}

Often when one collects images in experiments, each image is taken in a different orientation. 
%
Therefore, the first task in image analysis is to align or {\em register} the set of images. 
%
There are many ways to align images. 
%
``Template--based'' alignment \cite{ahuja2007template} is one technique, where one selects a specific image or {\em template}, and then tries to optimally align each image to that template.
%
However, this method can result in many misalignemnts for noisy images, and can be problematic if a ``good'' template image is not known {\em a priori}. 
%
Instead, we will use a technique called angular synchronization\cite{singer2011angular} to align sets of images.
%
Angular synchronization first computes the pairwise alignments for every possible pair of images, effectively using each image as a template for every other image.
%
From this large collection of pairwise alignments, it then tries to find optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignment.
%
Because we consider {\em all} pairwise alignments, angular synchronization is often much more robust to noise.

We will first discuss angular synchronization {\em only} considering rotations. 
%
Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations.
%
We assume that each signal $x_i$ is a noisy rotated copy of the underlying signal $x_{true}$ (which we are {\em not} given), such that 
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
%
Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$ so that we can align $x_1, \dots, x_m$. 
%
Up to noise, 
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
 note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\em estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$, 
i.e., $\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}

@@

To illustrate how angular synchronization works, we have converted the two--dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section). 
%
Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one--dimensional concentration profile. 
%
The images are not aligned, and so rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one image.
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization. 
%
The peaks are now (mostly) aligned.

\begin{figure}
\begin{subfigure}{0.17\textwidth}
\includegraphics[width=\textwidth]{illustrate_1d}
\caption{}
\label{subfig:1d_image_example}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{unregistered_unordered_1d}
\caption{}
\label{subfig:1d_unaligned_unordered}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{registered_unordered_1d}
\caption{}
\label{subfig:1d_aligned_unordered}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{registered_ordered_1d}
\caption{}
\label{subfig:1d_aligned_ordered}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{registered_ordered_vdm_1d}
\caption{}
\label{subfig:1d_aligned_ordered_vdm}
\end{subfigure}
\caption{}
\label{fig:1d_demo}
\end{figure}

\subsection{Ordering Using Diffusion Maps}

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
%
We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
%
DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold.
%
In our analysis, we will assume that our data lie on a one--dimensional curve or manifold, and that this one dimension is correlated with time. 
%
We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.

To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
%
We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
%
We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$,
\begin{equation}
W = 
\begin{bmatrix}
	1 & 1/2 & 0 & 0 & \dots & 0 \\
	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
	0 & 0 & 0 & \dots & 1 & 1/2 \\
	0 & 0 & 0 & \dots & 1/2 & 1 
\end{bmatrix}.
\end{equation}
%
$W_{ij}$ is large if $x_i$ is close to $x_j$,
%
If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \cdots \cos((m-1) \pi / m) \cos(\pi)]^T$.
%
Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data, i.e. $\phi_2(1) > \phi_2(2) > \dots > \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
%
Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.

@@

We demonstrate how we can apply these techniques to our (aligned)one--dimensional concentration profiles. 
%
We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our distances in \eqref{eq:dmaps_W}.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non--trivial) dmaps component, $\phi_2$.

\subsection{Vector Diffusion Maps}

Many times, we would to both align and order our data.
%
Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
We can combine the two steps into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
%
This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
%
It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.

@@

Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
%
The data has been ordered by $\psi_{1, 3}$.

\subsection{Registering Images with Respect to Translations and Rotations} \label{subsec:trans_rot_register}

We are not only interested in registering the one--dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two--dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations. 
%
However, for \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
%
The typical matrix representation for the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property. 
%
However, we can (approximately) represent rotations and translations in two--dimensions using three--dimensional rotation matrices.
%
We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere (see Figure \ref{fig:SO3_picture}).
%
Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions \footnote{Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.}.
%
These rotation matrices are (by definition) orthoganol, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
%
See the Appendix for more details.

\section{Results}

\begin{figure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{registered_ordered_vdm_2d}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{rank_corr_vdm}
\caption{}
\end{subfigure}
\end{figure}

\appendix

\section{Computing Optimal Alignments for One--Dimensional Signals}

Each signal shown in Figure \ref{fig:1d_demo} is discretized into 100 points. 
%
To compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.

\section{Computing Optimal Pairwise Alignments for Two--Dimensional Images}

@@

\section{Converting from $ISO(2)$ to $SO(3)$}

\begin{figure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_1}
\includegraphics[width=\textwidth]{sphere2_1}
\caption{}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_2}
\includegraphics[width=\textwidth]{sphere2_2}
\caption{}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_3}
\includegraphics[width=\textwidth]{sphere2_3}
\caption{}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_4}
\includegraphics[width=\textwidth]{sphere2_4}
\caption{}
\end{subfigure}
\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions. (a) The original image. (b) Rotation around the x--axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image. (c) Rotation around the y--axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation. (d) Rotation around the z--axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation. The top row shows the three--dimensional spheres, and the bottom row shows the (two--dimensional) surface of the sphere in which we are interested.}
\label{fig:SO3_picture}
\end{figure}

@@

\section{Computing Optimal Translations and Rotations from the Eigenvectors of $H$}

From the three--dimensional rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
we build the matrix $H$ and compute the estimates $R_{i, est}$ as described in Section \ref{subsec:angsynch}.
%
From the matrices $R_{1, est}, \dots, R_{m, est}$, we now wish to find the corresponding translations and rotations of the images.
%
Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
\begin{equation}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) \\
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) \\
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta) 
\end{aligned}
\end{equation}
%
We can then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
%
We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.

\bibliographystyle{plain}
\bibliography{../../references/references}

>>>>

%% PNAStmpl.tex
%% Template file to use for PNAS articles prepared in LaTeX
%% Version: Apr 14, 2008

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BASIC CLASS FILE 
%% PNAStwo for two column articles is called by default.
%% Uncomment PNASone for single column articles. One column class
%% and style files are available upon request from pnas@nas.edu.
%% (uncomment means get rid of the '%' in front of the command)

%\documentclass{pnasone}
\documentclass{pnastwo}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Changing position of text on physical page:
%% Since not all printers position
%% the printed page in the same place on the physical page,
%% you can change the position yourself here, if you need to:

% \advance\voffset -.5in % Minus dimension will raise the printed page on the 
                         %  physical page; positive dimension will lower it.

%% You may set the dimension to the size that you need.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL GRAPHICS STYLE FILE

%% Requires graphics style file (graphicx.sty), used for inserting
%% .eps files into LaTeX articles.
%% Note that inclusion of .eps files is for your reference only;
%% when submitting to PNAS please submit figures separately.

%% Type into the square brackets the name of the driver program 
%% that you are using. If you don't know, try dvips, which is the
%% most common PC driver, or textures for the Mac. These are the options:

% [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
% [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
% [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]

%\usepackage[dvips]{graphicx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL POSTSCRIPT FONT FILES

%% PostScript font files: You may need to edit the PNASoneF.sty
%% or PNAStwoF.sty file to make the font names match those on your system. 
%% Alternatively, you can leave the font style file commands commented out
%% and typeset your article using the default Computer Modern 
%% fonts (recommended). If accepted, your article will be typeset
%% at PNAS using PostScript fonts.

% Choose PNASoneF for one column; PNAStwoF for two column:
%\usepackage{PNASoneF}
%\usepackage{PNAStwoF}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ADDITIONAL OPTIONAL STYLE FILES

%% The AMS math files are commonly used to gain access to useful features
%% like extended math fonts and math commands.

\usepackage{amssymb,amsfonts,amsmath}

\usepackage{subcaption}
\graphicspath{ {paper_figures/} }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL MACRO FILES
%% Insert self-defined macros here.
%% \newcommand definitions are recommended; \def definitions are supported

%\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
%\def\s{\sigma}

\DeclareMathOperator*{\argmin}{arg\,min}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Don't type in anything in the following section:
%%%%%%%%%%%%
%% For PNAS Only:
\contributor{Submitted to Proceedings
of the National Academy of Sciences of the United States of America}
\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\copyrightyear{2008}
\issuedate{Issue Date}
\volume{Volume}
\issuenumber{Issue Number}
%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% For titles, only capitalize the first letter
%% \title{Almost sharp fronts for the surface quasi-geostrophic equation}

\title{Temporal ordering and registration of imaging data in cross--sectional studies of development}

%% Enter authors via the \author command.  
%% Use \affil to define affiliations.
%% (Leave no spaces between author name and \affil command)

%% Note that the \thanks{} command has been disabled in favor of
%% a generic, reserved space for PNAS publication footnotes.

%% \author{<author name>
%% \affil{<number>}{<Institution>}} One number for each institution.
%% The same number should be used for authors that
%% are affiliated with the same institution, after the first time
%% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
%% Then, before last author ...
%% \and
%% \author{<author name>
%% \affil{<number>}{}}

%% For example, assuming Garcia and Sonnery are both affiliated with
%% Universidad de Murcia:
%% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
%% United Kingdom},
%% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
%% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}

\author{Carmeline J. Dsilva\affil{1}{Departmnent of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}, 
Bomyi Lim\affil{1}{}, Stanislav Y. Shvartsman\affil{1}{} \affil{2}{Lewis--Seigler Instittue for Integrative Genomics, Princeton University, Princeton, NJ 08544}, \and Ioannis G. Kevrekidis\affil{1}{} \affil{3}{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ 08544}}

\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

%% The \maketitle command is necessary to build the title page.
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{article}

\begin{abstract} 
In studies of development, researchers are often presented with cross--sectional data, where each data point is a sample from a population fixed at a slightly different developmental time. 
%
The goal is then to temporally order the data to reconstruct the developmental dynamics. 
%
If each data point is a two-dimensional image, the images must first be aligned or registered before they can be temporally ordered. 
%
When such data sets are large, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand. 
%
We present an automatic way to register and temporally order cross--sectional data sets of images. We use relatively new mathematical techniques (angular synchronization for image registration, and diffusion maps for temporal ordering) which are applicable to a wide variety of data sets and require little {\em a priori} knowledge of the image features or the developmental dynamics. 
%
We demonstrate the utility of our methods using a collection of images from a study of {\em Drosophila} embryogenesis.
\end{abstract}

%% When adding keywords, separate each term with a straight line: |
\keywords{temporal ordering | image registration}

%% Optional for entering abbreviations, separate the abbreviation from
%% its definition with a comma, separate each pair with a semicolon:
%% for example:
%% \abbreviations{SAM, self-assembled monolayer; OTS,
%% octadecyltrichlorosilane}

% \abbreviations{}

%% The first letter of the article should be drop cap: \dropcap{}
%\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts

%% Enter the text of your article beginning here and ending before
%% \begin{acknowledgements}
%% Section head commands for your reference:
%% \section{}
%% \subsection{}
%% \subsubsection{}

\begin{itemize}
\item Want to study development
\item Often difficult/impossible to obtain longitudinal data
\item Instead, we are presented with cross--sectional data, and we need to order it to look at dynamics
\item This can be done by hand for small data sets (Figure \ref{fig:fish})
\item large data sets-- need automation (Figure \ref{fig:fluorescent_images})
\item we will present mathematical algorithms to register images and temporally order them
\item Use {\em Drosophila} as a case study
\end{itemize}

\section{Results and Discussion}

Often when one collects images in experiments, each image is taken in a different orientation. 
%
Therefore, the first task in image analysis is to align or {\em register} the set of images. 
%
There are many methods for image registration. 
%
``Template--based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then tries to optimally register each image to that template.
%
However, this method can result in many errors for noisy images, and can be problematic if a ``good'' template image is not known {\em a priori}. 
%
Instead, we will use a technique called angular synchronization\cite{singer2011angular} to register sets of images.
%
Angular synchronization first computes the pairwise alignments for every possible pair of images, effectively using each image as a template for every other image.
%
From this large collection of pairwise alignments, it then tries to find optimal alignment for each image, such that these individual alignments are most consistent (in some metric) with the pairwise alignments.
%
Because we consider {\em all} pairwise alignments, angular synchronization is often much more robust to noise.

We will first discuss angular synchronization {\em only} considering rotations. 
%
We have converted the two--dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section).
%
Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one--dimensional concentration profile; the profile is discretized into 100 points. 
%
Rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one image.
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization;
the peaks are aligned, even though we input no information about the existence or location of these peaks into our alogrithm.
%
We would like to note that, because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered {\em up to} a global rotation, and we are permitted rotate every point by a constant rotation.

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
%
We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
%
DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold.
%
In our analysis, we will assume that our data lie on a one--dimensional curve/ manifold, and that this one dimension is correlated with time. 
%
Therefore, sorting the data by the value of DMAPS coordinate will order the data along this one--dimensional curve, and will therefore temporally order the data.

We demonstrate how we can apply these techniques to our (aligned) one--dimensional concentration profiles. 
%
We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances for DMAPS (in \eqref{eq:dmaps_W}).
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non--trivial) DMAPS component, $\phi_2$. 
%
As expected, the intensity of the dpERK peaks grows as a function of time. 
%
Although DMAPS can recover a good ordering for the data, it cannot tell us about the absolute time of the data; we therefore have no information about the speed of this trajectory.
%
Furthermore, the direction of the trajectory is also unknown; determining which ``end'' of the DMAPS trajectory corresponds to the young embryos, and which ``end'' corresponds to the old embryos is a task that needs to be done by hand.

In many cases, we are interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
%
This technique is called vector diffusion maps (VDM) \cite{singer2012vector}.
%
It not only reduces the number of required eigencomputations, but also is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of aligning and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
%
The data has been ordered by $\psi_{1, 3}$, the first (non--trivial) VDM coordinate.

The same methodologies (with small modifications) can be applied to two--dimensional images.
%
Instead of having $100$--dimensional concentration vectors, we now work with images containing $10000$ pixels.
%
The two--dimensional images are registered with respect to all possible translations and rotations. 
%
We used vector diffusion maps to align and order the images shown in Figure~\ref{subfig:images_scrambled}.
%
The results are shown in Figure~\ref{subfig:images_vdm}.
%
The peaks are now aligned, and the intensity of dpERK grows as a function of time, as expected.

In these {\em Drosophila} experiments, we have the luxury of an independent time marker from membranes.
%
We can therefore evaluate the quality of our orderings obtained from VDM by comparing with the orderings obtained from the membrane markers.
%
The rank of each data point, computed using VDM, versus the rank of each data point computed from the membrane thickness, is shown in Figure~\ref{subfig:rank_corr}.
%
The rank correlation is 0.8665, indicating that our ordering from VDM is consistent with the ordering obtained from the membrane thickness.
%
There is more scatter for the younger embryos; this is to be expected, as the dpERK signal is very low in this range.

Although our methods are fairly robust, they do require a sufficient amount of data to ``see'' a smooth developmental trajectory.
%
We evaluated how much data is required by subsampling our data set (with replacement), and for each subsample, computing the temporal ordering using VDM and calculating the rank correlation between this ordering and the ordering using the membrane markers.
%
The results are shown in Figure~\ref{subfig:bootstrap}.
%
For small data sets, the average rank correlation is very low, indicating that the orderings from VDM are inaccurate. 
%
However, for moderately sized data sets (approximately 30 data points), the 
rank correlation begins to plateau and VDM can accurately order the data.

\section{Conclusions}

\begin{itemize}

\item Techniques for registration and ordering
\item Very general-- requires little knowledge of the types of images
\item Ordering-- different than other studies (TSP, MST, etc.)
\item assumes one--dimensionality of data, and one dimension is correlated with time
\item gives no sense of absolute time (speed) or space (where the ``zero point'' is)
\end{itemize}

%% == end of paper:

%% Optional Materials and Methods Section
%% The Materials and Methods section header will be added automatically.

%% Enter any subheads and the Materials and Methods text below.

\begin{materials}

\section{Angular synchronization}

Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations.
%
We assume that each signal $x_i$ is a noisy rotated copy of the underlying signal $x_{true}$ (which we are {\em not} given), such that 
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
%
Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$ so that we can align $x_1, \dots, x_m$. 
%
Up to noise, 
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
 note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\em estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$, 
i.e., $\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}
%
For the one--dimensional profiles shown in Figure~\ref{fig:1d_demo}, to compute the optimal angles in \eqref{eq:opt_angle}, we exhaustively search over all 100 possible shifts/rotations of the signals.

@@

\section{Diffusion Maps}

We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.
%
To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
%
We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
%
We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$,
\begin{equation}
W = 
\begin{bmatrix}
	1 & 1/2 & 0 & 0 & \dots & 0 \\
	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
	0 & 0 & 0 & \dots & 1 & 1/2 \\
	0 & 0 & 0 & \dots & 1/2 & 1 
\end{bmatrix}.
\end{equation}
%
$W_{ij}$ is large if $x_i$ is close to $x_j$,
%
If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \cdots \cos((m-1) \pi / m) \cos(\pi)]^T$.
%
Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data, i.e. $\phi_2(1) > \phi_2(2) > \dots > \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
%
Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.

@@

\section{Vector Diffusion Maps}

@@

\section{Registering Images with Respect to Translations and Rotations} \label{subsec:trans_rot_register}

We are not only interested in registering the one--dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two--dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations. 
%
However, for \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
%
The typical matrix representation for the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property. 
%
However, we can (approximately) represent rotations and translations in two--dimensions using three--dimensional rotation matrices.
%
We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere (see Figure \ref{fig:SO3_picture}).
%
Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions \footnote{Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.}.
%
These rotation matrices are (by definition) orthoganol, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.

@@

@@

From the three--dimensional rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
we build the matrix $H$ and compute the estimates $R_{i, est}$ as described in Section \ref{subsec:angsynch}.
%
From the matrices $R_{1, est}, \dots, R_{m, est}$, we now wish to find the corresponding translations and rotations of the images.
%
Because we are permitted to adjust our rotations by a global rotations, we first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
\begin{equation}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) \\
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) \\
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta) 
\end{aligned}
\end{equation}
%
We can then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image from the Euler angles.
%
We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.  
\end{materials}

%% Optional Appendix or Appendices
%% \appendix Appendix text...
%% or, for appendix with title, use square brackets:
%% \appendix[Appendix Title]

\begin{acknowledgments}
-- text of acknowledgments here, including grant info --
\end{acknowledgments}

%% PNAS does not support submission of supporting .tex files such as BibTeX.
%% Instead all references must be included in the article .tex document. 
%% If you currently use BibTeX, your bibliography is formed because the 
%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
%% .tex document. To conform to PNAS requirements, copy the reference listings
%% from your .bbl file and add them to the article .tex file, using the
%% bibliography environment described above.

%%  Contact pnas@nas.edu if you need assistance with your
%%  bibliography.

% Sample bibliography item in PNAS format:
%% \bibitem{in-text reference} comma-separated author names up to 5,
%% for more than 5 authors use first author last name et al. (year published)
%% article title  {\it Journal Name} volume #: start page-end page.
%% ie,
% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991) 
% A short C-terminal sequence is necessary and sufficient for the
% targeting of chitinases to the plant vacuole. 
% {\it Proc Natl Acad Sci USA} 88:10362-10366.

%% Enter the largest bibliography number in the facing curly brackets
%% following \begin{thebibliography}

\bibliographystyle{pnas}
\bibliography{../../references/references}

%\begin{thebibliography}{10}

%\bibitem{ahuja2007template}
%Sunil Ahuja, Ioannis~G Kevrekidis, and Clarence~W Rowley.
%\newblock Template-based stabilization of relative equilibria in systems with
%  continuous symmetry.
%\newblock {\em Journal of Nonlinear Science}, 17(2):109--143, 2007.
